---
title: "HARTWELL Data Validation"
author: "KIDS26-Team11"
date: "`r Sys.Date()`"
output: html_document
---

# Purpose

This notebook validates a TSV or CSV file against `validation_rules_schema_HARTWELL.json`, a plain `type`/`required`/`allowed` schema, and writes two outputs:

- rows that pass every check, ready to upload to LabKey
- rows that fail, with the specific field and reason for each failure

It performs validation only. Merging, standardization, and LabKey upload are handled elsewhere.

# Required packages

```{r install-packages, eval=FALSE}
install.packages(c("tidyverse", "jsonlite"))
```

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(jsonlite)
library(tools)
```

# User settings

```{r user-settings}
input_path  <- "../../analyses/HARTWELL_database_Master_MERGED.tsv"
schema_path <- "validation_rules_schema_HARTWELL.json"
out_prefix  <- "HARTWELL_database_Master_MERGED"
```

# Read the input file

Reads as CSV or TSV based on the file extension. All columns are read as text so blanks and formatting are preserved exactly as validation sees them.

```{r read-input}
read_table <- if (file_ext(input_path) == "csv") read_csv else read_tsv
data <- read_table(input_path, col_types = cols(.default = "c")) %>%
  mutate(row_id = row_number())

dim(data)
names(data)
```

# Read the schema

```{r read-schema}
rules <- fromJSON(schema_path, simplifyVector = FALSE)$hartwell_sample_rules

length(rules)
names(rules)
```

# Match schema fields to columns

A few schema field names don't match the raw HARTWELL column names (e.g. schema field `Lab_ID` vs. actual column `Lab ID #`). Those aliases are listed here rather than in the schema itself, so the schema stays plain `type`/`required`/`allowed`.

```{r field-aliases}
raw_column_aliases <- c(
  Lab_ID = "Lab ID #",
  Sample_Name = "Sample Name",
  SRM_Order_ID = "SRM Order #",
  SRM_Sample_ID = "SRM Sample #",
  Order_by = "Ordered by",
  Parental_Model_ID = "parental_model_id"
)

fields <- map(names(rules), function(field_name) {
  r <- rules[[field_name]]
  column <- case_when(
    field_name %in% names(data) ~ field_name,
    field_name %in% names(raw_column_aliases) ~ unname(raw_column_aliases[field_name]),
    TRUE ~ NA_character_
  )
  list(field = field_name, column = column, required = isTRUE(r$required),
       type = r$type, allowed = unlist(r$allowed))
})
```

# Check each field

For every schema field, checks:

- **required**: value must not be blank
- **allowed**: value, if present, must be in the schema's allowed list
- **boolean**: value, if present, must parse as `TRUE`/`FALSE`

Produces one row per (row, field, reason) failure.

```{r validate}
errors <- map_dfr(fields, function(f) {
  if (is.na(f$column)) {
    if (f$required) return(tibble(row_id = data$row_id, field = f$field, reason = "required column missing from input"))
    return(tibble())
  }

  values <- data[[f$column]]
  blank  <- is.na(values) | str_trim(values) == ""

  bind_rows(
    if (f$required) tibble(row_id = data$row_id[blank], field = f$field, reason = "required but blank"),
    if (!is.null(f$allowed)) {
      bad <- !blank & !(values %in% f$allowed)
      tibble(row_id = data$row_id[bad], field = f$field, reason = str_c("value '", values[bad], "' not allowed"))
    },
    if (identical(f$type, "boolean")) {
      bad <- !blank & !(str_to_lower(str_trim(values)) %in% c("true", "false", "t", "f", "1", "0"))
      tibble(row_id = data$row_id[bad], field = f$field, reason = str_c("'", values[bad], "' is not a valid boolean"))
    }
  )
})
```

# Split into valid and failing rows

```{r split}
row_errors <- errors %>%
  group_by(row_id) %>%
  summarise(validation_errors = str_c(field, ": ", reason, collapse = "; "), .groups = "drop")

data <- left_join(data, row_errors, by = "row_id")
valid_rows   <- data %>% filter(is.na(validation_errors)) %>% select(-row_id, -validation_errors)
invalid_rows <- data %>% filter(!is.na(validation_errors)) %>% select(-row_id)
```

# Most common validation failures

```{r failure-summary}
if (nrow(invalid_rows) > 0) {
  reasons <- str_split(invalid_rows$validation_errors, "; ") %>% unlist()
  field_only <- str_extract(reasons, "^[^:]+")
  sort(table(field_only), decreasing = TRUE)
}
```

# Export the outputs

```{r export}
write_tsv(valid_rows, str_c(out_prefix, "_valid.tsv"), na = "")
write_tsv(invalid_rows, str_c(out_prefix, "_validation_errors.tsv"), na = "")

cat("Valid output:", str_c(out_prefix, "_valid.tsv"), "\n")
cat("Errors output:", str_c(out_prefix, "_validation_errors.tsv"), "\n")
```

# Validation summary

```{r summary}
cat("Total rows:", nrow(data), "\n",
    "Valid rows:", nrow(valid_rows), "\n",
    "Failing rows:", nrow(invalid_rows), "\n")
```

# The notebook assumes:

- `input_path` is already merged/standardized data, one record per row.
- Column names in `input_path` match the schema's field names exactly, or are covered by `raw_column_aliases`.
- The schema file uses the same `type`/`required`/`allowed`/`boolean` structure as `validation_rules_schema_HARTWELL.json`.

[^1]: Developed as part of the St. Jude BioHackathon 2026. (Team-11)
