---
title: "Excel Metadata Ingestion and Merging"
author: "Asha Jacob Jannu"
date: "2026-09-16"
output: html_document
---

# Purpose

This notebook ingests metadata from multiple Excel workbooks and all tabs within each workbook, merges them into one dataframe, and exports the result as a CSV file.

It:
- Finds all `.xlsx` and `.xls` files in an input folder.
- Reads every sheet from every workbook.
- Reads all columns as text to avoid type conflicts across files.
- Preserves free-text fields such as `Notes`.
- Cleans column names.
- Adds `source_file` and `source_sheet`.
- Combines tables even when columns differ.
- Keeps missing values as `NA` in the final CSV.

This notebook performs data ingestion only. Schema normalization, value normalization, and validation should be handled separately.

# Required packages

Install these packages once if needed:

```{r install-packages, eval=FALSE}
install.packages(c("readxl", "dplyr", "purrr", "janitor", "readr"))
```

Load the packages:

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(purrr)
library(janitor)
library(readr)
```

# User settings

Set the folder containing the Excel files and the output CSV filename.

```{r user-settings}
input_dir <- "metadata/"
output_file <- "combined_metadata.csv"
```

# Find Excel files

```{r find-files}
files <- list.files(
  path = input_dir,
  pattern = "\\.(xlsx|xls|txt)$",
  full.names = TRUE,
  ignore.case = TRUE
)

if (length(files) == 0) {
  stop("No files were found in: ", input_dir)
}

cat("Excel files found:", length(files), "\n")
basename(files)
```

# Parse and merge input files

Here, each excel file is checked for sheet names. Every sheet is read separately and all columns are imported as text.

Reading all columns as text prevents type conflicts when the same column contains dates, numbers, IDs, or text across different files.

```{r parse-and-merge}
# Parse and merge .xls/.xlsx and .txt files
combined_data <- map_dfr(files, function(file) {

  if (grepl("\\.xlsx?$", file, ignore.case = TRUE)) {

    sheets <- excel_sheets(file)

    map_dfr(sheets, function(sheet) {
      df <- read_excel(path = file, sheet = sheet, col_types = "text") %>% clean_names()
      
      message(basename(file), " | ", sheet, " | Rows: ", nrow(df), " | Columns: ", ncol(df))
      
      df %>% mutate(source_file = basename(file), source_sheet = sheet)
    })

  } else if (grepl("\\.txt$", file, ignore.case = TRUE)) {

    df <- read_tsv(file, col_types = cols(.default = col_character())) %>% clean_names()
    
    message(basename(file), " | Rows: ", nrow(df), " | Columns: ", ncol(df))

    df %>% mutate(source_file = basename(file), source_sheet = NA_character_)
  }
})

# Remove completely empty columns
combined_data <- combined_data %>%
  select(where(~ !all(is.na(.x) | trimws(.x) == ""))) %>% as.data.frame(combined_data)

# Rename remaining unnamed columns
unnamed_cols <- names(combined_data)[grepl("^x[0-9]+$", names(combined_data))]

names(combined_data)[match(unnamed_cols, names(combined_data))] <- paste0("extra_unnamed_col", seq_along(unnamed_cols))

```

# Different columns across sheets

The sheets do not need to contain identical columns. For example, one sheet/file may contain:

```text
sample_id | species | tissue | notes
```

and another sheet/file:

```text
sample_id | assay | platform
```

The combined dataframe will contain the union of all columns:

```text
sample_id | species | tissue | notes | assay | platform
```
Columns missing from a particular sheet/file will contain `NA`.

# Free-text fields

Text containing spaces and commas remains in a single field. For example:

```text
Sample received on dry ice and sample quality was acceptable
```
remains one value in the `notes` column.

# Inspect the combined dataframe

```{r inspect-data}
dim(combined_data)
head(combined_data)
names(combined_data)
```

# Review source information

`source_file` contains the original Excel filename and `source_sheet` contains the original sheet name.

```{r inspect-source}
combined_data %>%
  count(source_file, source_sheet, name = "rows")
```

# Export the combined CSV

Missing values are written explicitly as `NA`.

```{r export-csv}
write_csv(combined_data, output_file, na = "NA")

cat("Combined CSV written to:", output_file, "\n")
```

# Ingestion summary

```{r ingestion-summary}
cat("Excel files processed:", length(files), "\n",
  "Rows in combined dataset:", nrow(combined_data), "\n",
  "Columns in combined dataset:", ncol(combined_data), "\n",
  "Output file:", output_file, "\n")
```


# The script assumes:

The parser assumes:
- The first row of each sheet contains the column headers.
- Each sheet contains one main table.
- Different sheets may contain different columns.
- Column equivalence is handled later during schema normalization.


[^1]: Developed as part of the St. Jude BioHackathon 2026. (Team-11)
